\addcontentsline{toc}{chapter}{Acknowledgments}
\thispagestyle{empty}

\vspace*{20mm}

\begin{center}
{\usekomafont{section} Acknowledgments}
\end{center}

\vspace{10mm}
First and most importantly, I want to thank my advisor Philipp Holl. In the past four months, I could always count on his assistance and valuable feedback. The weekly meetings helped me a lot; if I had problems at home, I usually got a detailed answer via mail within minutes. Without him, this thesis would not have been possible. Again, thank you very much Philipp for your support, dedication and understanding in the last few months! \\\\
I would also like to thank my supervisor Prof. Dr. Nils Thürey who gave me the opportunity to write this thesis at the Chair of Computer Graphics and Visualization. Many thanks for the personal meetings and interesting conversations. I have learned an incredible amount about GPU programming in this short time, which I will continue to utilize beyond this thesis.\\\\
Before I started my studies at TUM, I already knew that my thesis was supposed to be in the field of physics simulations. Therefore, I would like to thank my math and physics teacher Dr. Clemens Koerdt, who introduced us to numerical n-body gravitation simulations in addition to his lessons, which aroused my interest in this topic.\\\\
I also want to thank my Swabian fellow students and friends Christian, Yannick, Tobias and Karl for making the last 4.5 years in Munich very enjoyable! \textit{Thanks} you.\\\\
Last but not least I have to thank my family for the encouragement through my studies. None of this would have been possible without them. I especially want to thank my dad and brother who motivated and supported me in a very challenging phase of my life during my studies.
\cleardoublepage{}
